Litrepl: Literate Paper Processor Promoting Transparency More Than Reproducibility

### 2025-01-02 paper/paper.tex
Update paper

\usepackage[utf8]{inputenc}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{geometry}
\usepackage{hyperref}
\usepackage{fancyhdr}
\usepackage{svg}
\usepackage{comment}
\usepackage{float}
\usepackage[font=small,skip=0pt]{caption}

@@

\fancypagestyle{plain}{
  \fancyhf{}
  \lhead{Sergei Mironov}
  \rhead{Litrepl}
  \rfoot{\thepage}
}

@@

The concept of \textit{Literate Programming} was formulated by Donald Knuth,
suggesting a shift in focus from writing code to explaining to human beings what
we want a computer to do (\textcite{Knuth1984lp}). This approach is embodied in
the WEB system and its derived family of tools, which translate combined code
and narrative documents into human-readable documentation and machine-executable
code. Over time, the concept has evolved, showing a trend towards simplification
(\textcite{Ramsey1994lps}).

@@

A significant milestone in this area became the IPython interpreter
(\textcite{Perez2007IPython}), which later evolved into the Jupyter Project. The
creators introduced a new document format called the Jupyter Notebook
(\textcite{Kluyver2016jupnb}), characterized by a series of logical sections of
various types, including text and code, which could directly interact with
programming language interpreters. This interactive communication, akin to REPL
style programming, allows the creation of well-structured documents suitable for
presentations and sharing. The concept underpinning these developments is termed
\textit{Literate Computing}\cite{Perez2015blog}, which includes goals of
spanning a wide audience range, boosting reproducibility, and fostering
collaboration. To achieve these objectives, several technical decisions were
made, notably the introduction of bidirectional communication between the
computational core, known as the Jupyter Kernel, and the Notebook web-based
renderer, along with another layer of client-server communication between the
renderer and the user’s web browser.

While we recognize the importance of all goals within the Literate Computing
framework, we argue that the goal of reproducibility is paramount. Successfully
addressing reproducibility would alone suffice to enhance communication over the
time and space separated researchers and significantly expand the audience.
However, as it became clear (\textcite{Dolstra2010}), this challenge extends
beyond the scope of a single human-computer interaction system, and even beyond
the typical boundaries of software distribution management for a particular
programming language. A comprehensive solution to the software deployment
problem should operate at the entire operating system level.

Following \textcite{Vallet2022}, we suggest changing the focus of human-computer
interaction towards simplicity and transparency. We saw an opportunity to
implement a tool that would offer familiar REPL-style editing, be compatible
with existing code editors and text formats, thus maintaining familiar editing
practices, and have a gradually smaller codebase.

We introduce \textit{Litrepl}, a processor designed to achieve these goals
through several means. First, we employ straightforward bidirectional text
streams for inter-process communication with language interpreters to evaluate
code.  Second, we advocate for the reuse of existing text document formats. In
both the Markdown and \Latex evaluators that we have implemented, simplified
parsers are utilized to distinguish code and result sections from the rest of
the document.  As of now, we support Python and Shell interpreter families, as
well as a custom large language model communication interpreter. Finally, we
strive to leverage POSIX\cite{POSIX2024} operating system facilities as much as
possible.

@@

Litrepl is implemented as a command-line text processor utility. Its primary
function is to take a text document as input, process it according to specified
command line arguments and environment settings, and then output the resulting
document through its standard output.

>>>>

\usepackage[utf8]{inputenc}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{geometry}
\usepackage{hyperref}
% \usepackage{fancyhdr}
\usepackage{svg}
\usepackage{comment}
\usepackage{float}
\usepackage[font=small,skip=0pt]{caption}

@@

% \fancypagestyle{plain}{
%   \fancyhf{}
%   \lhead{Sergei Mironov}
%   \rhead{Litrepl}
%   \rfoot{\thepage}
% }

@@

The concept of \textit{Literate Programming} was formulated by Donald Knuth,
suggesting a shift in focus from writing code to explaining to human beings what
we want a computer to do (\textcite{Knuth1984lp}). This approach is embodied in
the WEB system and its descendant family of tools, which translate combined code
and narrative documents into human-readable documentation and machine-executable
code. Over time, the concept has evolved, showing a trend towards simplification
(\textcite{Ramsey1994lps}).

@@

A significant milestone in this field was the IPython interpreter
(\textcite{Perez2007IPython}), which later evolved into the Jupyter Project. The
creators introduced a new document format called the Jupyter Notebook
(\textcite{Kluyver2016jupnb}), characterized by a series of logical sections of
various types, including text and code, which could directly interact with
programming language interpreters. This interactive communication, akin to REPL
style programming, allows the creation of well-structured documents suitable for
presentations and sharing. The concept underpinning these developments is termed
\textit{Literate Computing}\cite{Perez2015blog}, which includes goals of
spanning a wide audience range, boosting reproducibility, and fostering
collaboration. To achieve these objectives, several technical decisions were
made, notably the introduction of bidirectional communication between the
computational core, known as the Jupyter Kernel, and the Notebook web-based
renderer, along with another layer of client-server communication between the
web-server and the user’s web browser.

While we recognize the importance of all goals within the Literate Computing
framework, we argue that the goal of reproducibility is paramount. Successfully
addressing reproducibility would alone suffice to enhance communication among
time-separated and space-separated researchers and significantly expand the
audience. However, as it became clear (\textcite{Dolstra2010}), this challenge
extends beyond the scope of a single human-computer interaction system, and even
beyond the typical boundaries of software distribution management for a
particular programming language. A comprehensive solution to the software
deployment problem should operate at the entire operating system level.

Following \textcite{Vallet2022}, we suggest changing the focus of human-computer
interaction towards simplicity and transparency. We saw an opportunity to
implement a tool that would offer REPL-style editing, be compatible with
existing code editors and text formats, thus maintaining familiar editing
practices, and have a significantly smaller codebase.

We introduce \textit{Litrepl}, a text processor that employs the following
approaches: first, utilizing straightforward bidirectional text streams for
inter-process communication with language interpreters to evaluate code; second,
advocating for the reuse of existing text document formats. In both the Markdown
and \LaTeX\ evaluators we have implemented, simplified parsers are used to
distinguish code and result sections from the rest of the document. As of now,
we support Python and Shell interpreter families, as well as a custom large
language model communication interpreter. Finally, we strive to leverage
POSIX\cite{POSIX2024} operating system facilities as much as possible.

@@

Litrepl is implemented as a command-line text utility. Its primary function is
to take a text document as input, process it according to specified command-line
arguments and environment settings, and then output the resulting document
through its standard output.
